---
title: "The State of Global Health & Vulnerability as a result of Climate Change"
format: 
  revealjs: 
    embed-resources: true 
editor: visual 
---



```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(lubridate)
library(dplyr)
library(rvest)
library(magrittr)
library(ggmap)
library(stringr)
library(rworldmap)
library(scales)
library(plotly)
library(leaflet)
library(gganimate)

```

## Notre Dame Global Adaptation Dataset
- Data from 192 countries about the vulnerability of countries to climate-change challenges
- Vulnerability score from 0-1 is made up of multiple indicators from the health sector: 
  - Projected change of deaths from climate change induced diseases 
  - Access to improved sanitation facilities  
  - Number of medical staff (physicians, nurses and midwives) 


## Vulnerability Over Time

```{r,  echo=FALSE}
health_vul <- read_csv("./resources/vulnerability/health.csv")
health_vul <- health_vul %>% 
  pivot_longer(3:28, names_to="year", values_to="vun") 

Continent2.0 <- health_vul %>%
 mutate(Name = if_else(Name == "Congo, the Democratic Republic o", "DR Congo", Name)) %>%
  mutate(Name = if_else(Name == "Russian Federation", "Russia", Name)) %>%
  mutate(Name = if_else(Name == "United Kingdom", "U.K", Name)) %>%
  mutate(Name = if_else(Name == "Papua New Guinea", "P.N.G", Name)) %>%
  mutate(Name = if_else(Name == "Venezuela, Bolivarian Republic o", "Venezuela", Name)) %>%
  filter(Name%in%c("China","India", "Indonesia", "Pakistan", "Bangladesh","Russia","Germany", "U.K", "France", "Italy","Nigeria","Ethiopia", "Egypt", "DR Congo", "South Africa","United States","Mexico", "Canada", "Guatemala", "Cuba","Australia","P.N.G", "New Zealand", "Fiji", "Solomon Islands","Brazil","Colombia", "Argentina", "Peru", "Venezuela")) %>%
  mutate(Continent =
            case_when(
            Name%in%c("Brazil","Colombia", "Argentina", "Peru", "Venezuela") ~
              "South America",
            Name%in%c("Australia","P.N.G", "New Zealand", "Fiji", "Solomon Islands") ~ "Oceania",
            Name%in%c("United States","Mexico", "Canada", "Guatemala", "Cuba") ~ "North America",
            Name%in%c("Russia","Germany", "U.K", "France", "Italy") ~ "Europe",
            Name%in%c("Nigeria","Ethiopia", "Egypt", "DR Congo", "South Africa") ~ "Africa",
          Name%in%c("China","India", "Indonesia", "Pakistan", "Bangladesh") ~ "Asia"))



p <- (Continent2.0 %>%
  ggplot(aes(x = year, y = vun, color = Continent, group = Name,text = paste0('Name: ',Name,'\n','Year: ',year,'\nVulnerability: ',round(vun,2))))+
  geom_line()+
    scale_color_viridis_d(option='turbo') +
    #scale_color_manual(values = cbp1)+
    theme_classic()+
  theme(axis.text.x = element_text(angle = 45, hjust=1))+
labs( y = "vulnerability",color = 'Continent') +
  theme(panel.background = element_rect(fill = "white", color = "white"), legend.background = element_rect(fill = "white")) +
  theme(plot.background = element_rect(fill = "white", color = "light grey"), plot.margin = margin(15, 2, 2, 2, "pt"),)+
  theme(plot.title = element_text(face = "bold")))

p %>%
  ggplotly(tooltip = c('text')) %>%
layout(title = list(text = paste0("Global Health Vulnerability ","<br><sup>","Showing the change over time in different countires","</sup>")))
```

---

### Zooming In

```{r, echo = FALSE, height = 12}
Continent2.0$year <- as.numeric(as.character(Continent2.0$year))

p_facet <- (Continent2.0 %>%
  ggplot(aes(x = year, y = vun, color = Continent, group = Name,text = paste0('Name: ',Name,'\n','Year: ',year,'\nVulnerability: ',round(vun,2))))+
  geom_line()+
    scale_color_viridis_d(option='turbo') +
    #scale_color_manual(values = cbp1)+
    theme_classic()+
  theme(axis.text.x = element_text(angle = 45, hjust=1))+
labs( y = "vulnerability",color = 'Continent') +
  theme(panel.background = element_rect(fill = "white", color = "white"), legend.background = element_rect(fill = "white")) +
  theme(plot.background = element_rect(fill = "white", color = "light grey"), plot.margin = margin(15, 2, 2, 2, "pt"),)+
  theme(plot.title = element_text(face = "bold"))) +
  theme(axis.title.x = element_text(margin = margin(t = 8))) + facet_wrap(~Continent)

p_facet %>%
  ggplotly(tooltip = c('text')) %>%
layout(title = list(text = paste0("Global Health Vulnerability by Continent")))


```


```{r, echo=FALSE}
#Importing datasets

library(readr)
id_health_medstaff <- read_csv("resources/indicators/id_heal_04/input.csv", 
    col_types = cols(`1995` = col_double(), 
        `2015` = col_double()))
library(readr)
gdp <- read_csv("resources/indicators/gdp/score.csv", 
    col_types = cols(`1995` = col_skip(), 
        `1996` = col_skip(), `1997` = col_skip(), 
        `1998` = col_skip(), `1999` = col_skip(), 
        `2000` = col_skip(), `2001` = col_skip(), 
        `2002` = col_skip(), `2003` = col_skip(), 
        `2004` = col_skip(), `2005` = col_skip(), 
        `2006` = col_skip(), `2007` = col_skip(), 
        `2008` = col_skip(), `2009` = col_skip(), 
        `2010` = col_skip(), `2011` = col_skip(), 
        `2012` = col_skip(), `2013` = col_skip(), 
        `2014` = col_skip(), `2015` = col_skip(), 
        `2016` = col_skip(), `2017` = col_skip(), 
        `2018` = col_skip(), `2019` = col_skip()))


pop <- read_csv("resources/indicators/pop/input.csv", 
    col_types = cols(`1995` = col_skip(), 
        `1996` = col_skip(), `1997` = col_skip(), 
        `1998` = col_skip(), `1999` = col_skip(), 
        `2000` = col_skip(), `2001` = col_skip(), 
        `2002` = col_skip(), `2003` = col_skip(), 
        `2004` = col_skip(), `2005` = col_skip(), 
        `2006` = col_skip(), `2007` = col_skip(), 
        `2008` = col_skip(), `2009` = col_skip(), 
        `2010` = col_skip(), `2011` = col_skip(), 
        `2012` = col_skip(), `2013` = col_skip(), 
        `2014` = col_skip(), `2015` = col_skip(), 
        `2016` = col_skip(), `2017` = col_skip(), 
        `2018` = col_skip(), `2019` = col_skip()))


#projected percent increase in climate-change induced deaths
projecteddeath_perc <- read_csv("resources/indicators/id_heal_01/input.csv",
    col_types = cols(`1995` = col_skip(), 
        `1996` = col_skip(), `1997` = col_skip(), 
        `1998` = col_skip(), `1999` = col_skip(), 
        `2000` = col_skip(), `2001` = col_skip(), 
        `2002` = col_skip(), `2003` = col_skip(), 
        `2004` = col_skip(), `2005` = col_skip(), 
        `2006` = col_skip(), `2007` = col_skip(), 
        `2008` = col_skip(), `2009` = col_skip(), 
        `2010` = col_skip(), `2011` = col_skip(), 
        `2012` = col_skip(), `2013` = col_skip(), 
        `2014` = col_skip(), `2015` = col_skip(), 
        `2016` = col_skip(), `2017` = col_skip(), 
        `2018` = col_skip(), `2019` = col_skip()))

projecteddeath_perc_all <- read_csv("resources/indicators/id_heal_01/input.csv")

vulhealth <- read_csv("resources/vulnerability/health.csv")


```

```{r, echo=FALSE}

id_health_medstaff <- id_health_medstaff %>% 
  group_by(Name) %>% 
  mutate(percchange = .data[["2020"]] - .data[["1995"]]) %>% 
  arrange(desc(percchange))

#mean(id_health_medstaff$percchange,na.rm=T)
  

```

```{r, echo=FALSE}
pop2020 <- pop %>% 
   rename("pop2020" = "2020")

gdp2020 <- gdp %>% 
   rename("gdp2020" = "2020")

medstaff2020 <- id_health_medstaff %>% 
  rename("medstaff2020" = "2020") %>% 
  select(ISO3,Name,medstaff2020)

projecteddeathperc2020 <- projecteddeath_perc %>% 
  rename("projecteddeathperc2020" = "2020")

sanitationall <- read_csv("resources/indicators/id_heal_06/input.csv")

sanitation2020 <- sanitationall %>% 
  select(ISO3, Name, "2020") %>% 
  rename("sanitation2020" = "2020")


data2020 <- pop2020 %>% 
  left_join(gdp2020) %>% 
  left_join(medstaff2020) %>% 
  left_join(projecteddeathperc2020) %>% 
  left_join(sanitation2020)
  
```

```{r, echo=FALSE}

  hipopdata2020 <- data2020 %>% 
  arrange(desc(pop2020)) %>% 
  filter(pop2020 > 126000000)

  mostpopdata2020 <- data2020 %>% 
  arrange(desc(pop2020)) %>% 
  filter(pop2020 > 1000000)
```

```{r echo=FALSE}

#Will Not Include Top 10 country population Viz 
pdp <- ggplot(data = hipopdata2020, aes(x = gdp2020)) +
  geom_text(aes(x = gdp2020, y = pop2020, label = Name, color = projecteddeathperc2020), size=2.6) +
  geom_point(aes(x = gdp2020, y = pop2020, label = Name, color = projecteddeathperc2020, size = projecteddeathperc2020), alpha = .1) + 
  labs(title = "Projected % Increase in Climate-Change Induced Deaths \nBy GDP for the 10 Most Populous Countries", x = "GDP 2020 ($)", y = "Population 2020", color = "Projected Climate-change \nInduced Death % Increase") + guides(size="none") + scale_colour_gradient(low = 'blue', high = 'red') + theme_bw() + scale_y_continuous(limits = c(0, 1450000000), breaks=seq(0, 2000000000, by = 100000000), labels = scales::label_number_si()) +
  scale_x_continuous(limits = c(0, 70000), breaks=seq(0, 70000, by = 10000), labels = scales::label_number_si())

allpdp <- ggplot(data = data2020, aes(x = gdp2020)) +
  geom_point(aes(x = gdp2020, y = pop2020, label = Name, color = projecteddeathperc2020, size = pop2020), alpha = .2) + 
  labs(title = "Projected % Increase in Climate-Change Induced Deaths", x = "GDP 2020 ($)", y = "Population 2020", color = "Projected Climate-change \nInduced Death % Increase") + guides(size="none") + scale_colour_gradient(low = 'blue', high = 'red') + theme_bw() + scale_y_continuous(limits = c(0, 1450000000), breaks=seq(0, 2000000000, by = 100000000), labels = scales::label_number_si()) +
  scale_x_continuous(limits = c(0, 70000), breaks=seq(0, 70000, by = 10000), labels = scales::label_number_si())


```

---

### Climate-Change Induced Diseases

```{r echo=FALSE}
ggplotly(allpdp)

```

---

### Climate-Change Induced Diseases 

```{r,  echo=FALSE}
allpdp2 <- ggplot(data = data2020, aes(x = gdp2020)) +
  geom_point(aes(x = gdp2020, y = projecteddeathperc2020, label = Name, color = projecteddeathperc2020, size = pop2020), alpha = .2) + 
  labs(title = "Projected % Increase in Climate-Change Induced Deaths", x = "GDP 2020 ($)", y = "Projected Climate-change Induced Death % Increase", color = "%") + guides(size="none") + scale_colour_gradient(low = 'blue', high = 'red') + theme_bw() + scale_y_continuous(limits = c(1, 1.2), breaks=seq(1, 1.2, by = .02)) +
  scale_x_continuous(limits = c(0, 70000), breaks=seq(0, 70000, by = 10000), labels = scales::label_number_si())

ggplotly(allpdp2)
```

---

### Percent Change Over Time

```{r, echo = FALSE}





library(rnaturalearth)
map <- ne_countries()
names(map)[names(map) == "iso_a3"] <- "ISO3"
names(map)[names(map) == "name"] <- "NAME"

map$projecteddeathperc2020 <- data2020$projecteddeathperc2020[match(map$ISO3,data2020$ISO3)]

pal <- colorBin(
  palette = "magma", domain = map$projecteddeathperc2020,
  bins = seq(1, max(map$projecteddeathperc2020, na.rm = TRUE) + 0, by = .01)
)

map$labels <- paste0(
  "<strong> Country: </strong> ",
  map$NAME, "<br/> ",
  "<strong> projecteddeathperc2020: </strong> ",
  map$projecteddeathperc2020, "<br/> "
) %>%
  lapply(htmltools::HTML)

leafletdata2020 <- leaflet(map) %>%
  addTiles() %>%
  setView(lng = 0, lat = 30, zoom = 2) %>%
  addPolygons(
    fillColor = ~ pal(projecteddeathperc2020),
    color = "white",
    weight = 2,
    fillOpacity = 1,
    label = ~labels,
    highlight = highlightOptions(
      color = "black",
      bringToFront = TRUE
    )
  ) %>%
  leaflet::addLegend(
    pal = pal, values = ~projecteddeathperc2020,
    opacity = 0.7, title = "projecteddeathperc2020"
  )

leafletdata2020

```

## Sanitation Accessibility %

```{r, echo = FALSE}

map$sanitation2020 <- data2020$sanitation2020[match(map$ISO3,data2020$ISO3)]

pal <- colorNumeric(
  palette = "magma", domain = map$sanitation2020,
)


map$labels <- paste0(
  "<strong> Country: </strong> ",
  map$NAME, "<br/> ",
  "<strong> sanitation2020: </strong> ",
  map$sanitation2020, "<br/> "
) %>%
  lapply(htmltools::HTML)

leafletdata2020san <- leaflet(map) %>%
  addTiles() %>%
  setView(lng = 0, lat = 30, zoom = 2) %>%
  addPolygons(
    fillColor = ~ pal(sanitation2020),
    color = "white",
    weight = 2,
    fillOpacity = 1,
    label = ~labels,
    highlight = highlightOptions(
      color = "black",
      bringToFront = TRUE
    )
  ) %>%
  leaflet::addLegend(
    pal = pal, values = ~sanitation2020,
    opacity = 0.7, title = "sanitation2020"
  )

leafletdata2020san

```


```{r}

map$sanitation2020 <- data2020$sanitation2020[match(map$ISO3,data2020$ISO3)]

pal <- colorNumeric(
  palette = "magma", domain = map$sanitation2020,
)


map$labels <- paste0(
  "<strong> Country: </strong> ",
  map$NAME, "<br/> ",
  "<strong> sanitation2020: </strong> ",
  map$sanitation2020, "<br/> "
) %>%
  lapply(htmltools::HTML)

leafletdata2020san <- leaflet(map) %>%
  addTiles() %>%
  setView(lng = 0, lat = 30, zoom = 2) %>%
  addPolygons(
    fillColor = ~ pal(sanitation2020),
    color = "white",
    weight = 2,
    fillOpacity = 1,
    label = ~labels,
    highlight = highlightOptions(
      color = "black",
      bringToFront = TRUE
    )
  ) %>%
  leaflet::addLegend(
    pal = pal, values = ~sanitation2020,
    opacity = 0.7, title = "sanitation2020"
  )

leafletdata2020san

```


## Conclusion
- Vulnerability looks different for each country
  - Different factors, levels of change
- Highly correlation with GDP and level of development
- The countries most vulnerable are in the global south, compounded by multiple levels of vulnerability
- High variation of vulnerability over time
### Vulnerability is multi-faceted



